---
title: "Analyzing Data using R"
author: "By Shuhei Kitamura"
output:
  html_document: default
---

### Outline<a id='top'></a>
1. [Summary Tables](#sec1)
2. [Figures](#sec2)
3. [Regression Tables](#sec3)

```{r, eval=FALSE}
# load packages
library(texreg)
library(xtable)
library(lfe) 
library(tidyverse)
#library(plyr)
```

```{r, eval=FALSE}
# set the display options (not necessary)
options(repr.matrix.max.rows=200, repr.matrix.max.cols=100) 
options(repr.plot.width=4, repr.plot.height=4) 
```

```{r, eval=FALSE}
# set a seed
set.seed(123456789) 
```

```{r, eval=FALSE}
# set the working directory (if necessary)
# setwd('...') 
```

```{r, eval=FALSE}
# importing data
data <- read.table('data_use_R.csv', sep=",", header=TRUE) # import data 
print(nrow(data))
data = data[data['gelec_total']!=0, ] # drop rows
print(nrow(data))
```

## 1. Summary Tables<a id='sec1'></a>
- A summary table contains the statistics of your data such as mean, std, and the number of observation.
- Summary tables are very useful because people can understand the details of your data.
- To make a summary table, you can use `xtable` in `xtable` package.
- You can produce an output table in several format. I suggest the LaTeX format.
    - The reason: You can compile the table easily when you write a paper.
- For the moment, we produce files with the `txt` file extension.
    - If you already have TeX in your computer, change the file extension from `txt` to `tex`.
            
[back to top](#top)
```{r, eval=FALSE}
sum_table <- data %>%
    select('gelec_dem', 'gelec_rep', 'gelec_oth', 'gelec_total', 'rep_share', 'dem_share', 'elec_year', 'temp_mean', 'temp_max_max', 'temp_max_mean') %>%
    gather(key='key', value='value') %>%
    group_by(key) %>%
    summarize(mean = mean(value, na.rm=TRUE), 
              std = sd(value, na.rm=TRUE),
              min = min(value, na.rm=TRUE),
              max = max(value, na.rm=TRUE),
              n = sum(!is.na(value))) 

sum_table[['id']] <- c(6, 7, 1, 3, 2, 4, 5, 9, 10, 8) 
sum_table <- sum_table[order(sum_table[ ,'id']),][-7] # order sum_table
print(xtable(sum_table, digits=2), type="latex", file='sum_stat_R.txt')
```

## 2. Figures<a id='sec2'></a>
- Good papers always have figures that summarize results well.
    - You should not write a paper only with boring tables!
- A very useful package for plotting figures is `ggplot2`.
    - A [cheat sheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf) is available.  
- There are several types of figures: histograms, density plots, scatter plots, bar plots, line plots...
    - Each type has different purposes.
- Regarding the format, I suggest either `png` or `jpg`.

[back to top](#top)

### Histograms
- Histograms are often used to show the distributions of your data.
```{r, eval=FALSE}
ggplot(data) + 
    geom_histogram(aes(x=rep_share), breaks=c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0), alpha=0.3, fill='red', na.rm=TRUE)  +
    geom_histogram(aes(x=dem_share), breaks=c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 1.0), alpha=0.3, fill='blue', na.rm=TRUE)  +
    labs(title="Vote share, 2008-2014", x="") +
    theme(
        panel.background = element_rect(fill=NA),
        panel.border = element_rect(fill=NA, color='grey75'),
        axis.ticks = element_line(color='grey85'),
        legend.position = 'none',
        plot.title = element_text(hjust=0.5, size=9),
        axis.title = element_text(size=9),
        axis.text = element_text(size=9)
    )
ggsave('hist_rep_share_R.png', width=4, height=4) # save figure
```

### Density Plots
- Density plots are a variation of histograms.
    - For density plots, you employ some method to smooth the distribution.
    - It is not affected by, e.g., how you choose bins.
```{r, eval=FALSE}
data1 <- data.frame(value=rnorm(1000,0,1))
ggplot(data1, aes(x=value)) +
    geom_histogram(bins=100, alpha=0.3, aes(y=..density..)) +
    stat_density(geom="line", color='red') +
    labs(title="title", x="", y="") +
    theme(
        panel.background = element_rect(fill=NA),
        panel.border = element_rect(fill=NA, color='grey75'),
        axis.ticks = element_line(color='grey85'),
        legend.position = 'none',
        plot.title = element_text(hjust=0.5, size=9),
        axis.title = element_text(size=9),
        axis.text = element_text(size=9)
    )
```

### Bar Plots
- Bar plots are often used to compare statistics (e.g., mean) for different groups.
```{r, eval=FALSE}
data1 <- data.frame(
    name = c("tom", "jerry", "spike", "tyke"),
    height = c(1.75, 1.82, 1.65, 1.4),
    treatment = c(1, 1, 0, 0))

ggplot(data1, aes(x=treatment, y=height, fill=factor(treatment))) + 
    stat_summary(fun=mean, geom='bar', alpha=0.3) +
    scale_fill_manual(values=c('red', 'blue')) +
    scale_x_continuous(breaks=c(0,1)) +
    labs(title="", x="treatment", y="average height") +
    theme(
        panel.background = element_rect(fill=NA),
        panel.border = element_rect(fill=NA, color='grey75'),
        axis.ticks = element_line(color='grey85'),
        legend.position = 'none',
        plot.title = element_text(hjust=0.5, size=9),
        axis.title = element_text(size=9),
        axis.text = element_text(size=9)
    )
```

### Scatter Plots
- Scatter plots are often used to show a relationship between two samples.
```{r, eval=FALSE}
ggplot(data, aes(x=temp_max_max, y=rep_share)) +
    geom_point(color='blue', na.rm=TRUE) +
    labs(title="",x="Mean temperature",y="Republican vote share") +
    theme(
        panel.background = element_rect(fill = NA),
        panel.border = element_rect(fill = NA, color = "grey75"),
        axis.ticks = element_line(color = "grey85"),
        legend.position = "none",
        plot.title = element_text(hjust = 0.5, size=9),
        axis.title = element_text(size=9),
        axis.text = element_text(size=9)
    )
```

### Line Plots
- Line plots are often used to show time trends.
```{r, eval=FALSE}
n <- 1000
data1 <- data.frame(
    sample = 1:n,
    group=rep(c("one", "two", "three"), each=n),
    value=c(cumsum(rnorm(n,0,1)), cumsum(rnorm(n,0,1)), cumsum(rnorm(n,0,1))))

ggplot(data1, aes(x=sample, y=value)) +
    geom_line(aes(colour=group)) +
    labs(title="Title", x="", y="") +
    scale_color_manual(name="", values=c("blue","red","yellow")) +
    theme(
        panel.background = element_rect(fill=NA),
        panel.border = element_rect(fill=NA, color="grey75"),
        axis.ticks = element_line(color="grey85"),
        legend.position = "bottom",
        legend.key = element_blank(),
        legend.text = element_text(size=8),
        plot.title = element_text(hjust=0.5, size=9),
        axis.title = element_text(size=9),
        axis.text = element_text(size=9)
    )
```

## 3. Regression Tables<a id='sec3'></a>
- To run a regression, you can use, e.g., `lm` or `felm` in `lfe` package.
- To produce a table of the regression results, you can use `stargazer` in `stargazer` package.
    - If you want to produce the result in tex format, use `type=latex` and replace `.txt` with `.tex`.
- How can you interpret the regression results?

[back to top](#top)
```{r, eval=FALSE}
reg1 <- felm(rep_share ~ ln_temp_max_max, data=data)
reg2 <- felm(rep_share ~ ln_temp_max_max | state_short + elec_year, data=data)
reg3 <- felm(rep_share ~ ln_temp_max_max | state_short + elec_year | 0 | state_short, data=data)

texreg(list(reg1, reg2, reg3), 
       file='estimates_R.txt',
       custom.model.names = c("OLS", "FE", "FE+cluster"),
       caption="Table 1: Correlation between Election Day temperature and Senate Republican vote share",
       omit.coef="Intercept", digit=3)
```













