---
title: "Cleaning Data using R"
author: "By Shuhei Kitamura"
output:
  html_document: default
---


#### 　
- **Research Question**: Is there any relationship between Election-Day temperature and electoral outcomes?
- In this course, we use two datasets
    - [Senate election results](https://transition.fec.gov/pubrec/electionresults.shtml) in the United States
    - [Daily temperature](https://aqs.epa.gov/aqsweb/airdata/download_files.html)
- We start by cleaning the datasets.

### Outline<a id='top'></a>
1. [Importing Data](#sec1)
2. [Treating Missing Values](#sec2)
3. [Selecting Columns](#sec3)
4. [Selecting Rows](#sec4)
5. [Treating Other Values](#sec5)
6. [Saving Data](#sec6)

```{r, eval=FALSE}
# load packages
library(readxl) 
```

```{r, eval=FALSE}
# set the display options (not necessary)
options(repr.matrix.max.rows=200, repr.matrix.max.cols=100) 
```

```{r, eval=FALSE}
# set the working directory (if necessary)
# "\" mark does not work. use a forward slash "/" instead. 
#setwd('...') 
```

## 1. Importing Data<a id='sec1'></a>
- For tsv files, use `read.table()`.
- For csv files, use `read.table()`.
    - Be sure to include `sep=','`.
    - By adding `header=TRUE`, R recognizes that the first row is column names. 
- For excel files, use `read_excel()` in `readxl` package.
    - In that case, you need to convert the imported data to data frame using `as.data.frame()`.
    - Alternatively, use `read.xlsx()` in `xlsx` package or `readWorksheetFromFile()` in `XLConnect` package. Both packages seem to require 64-bit Java for Windows users.
- Alternatively, you could use `fread` in `data.table` package for large csv or tsv data.
- Datasets
    - `elec_senate.xlsx`: US Senate general election results 2008-2014
    - `daily_TEMP_XXXX.csv`: US temperature 2008-2014
    
[back to top](#top)
```{r, eval=FALSE}
elec_data <- list() # make an empty list
temp_data <- list()
for (year in seq(2008,2014,by=2)) {
    elec_data[[as.character(year)]] <- as.data.frame(read_excel('elec_senate.xlsx', sheet=as.character(year)))
    temp_data[[as.character(year)]] <- read.table(paste0('daily_TEMP_', year, '.csv'), header=TRUE, sep=',')
}
```

### Checking data entries
- What is the type of `elec_data[['2008']]` and `elec_data['2008']`? Use `class()`.
- Print the list of column names in `elec_data[['2008']]`. Hint: `colnames()`.
- To see samples of the data. Use `head(mydf, #)` or `tail(mydf, #)`, where `#` means the number of rows. 
```{r, eval=FALSE}
#
```

- Next, check whether data have unique and non-missing ids.
- Which variables are ids in `elec_data[['2008']]`?
    - To check the uniqueness, use `any(duplicated(mydf$column))`.
    - To check the non-missingness, use `any(is.na(mydf$column))`.
- You can also check the non-missingness using `summary()`.
- Check if ids are unique and non-missing in `elec_data[['2008']]`.
```{r, eval=FALSE}
#
```

## 2. Treating Missing Values<a id='sec2'></a>
- Print `elec_data[['2008']]`. We already know that some items are missing.
- Recall: There are several strategies to handle missing data:
    1. Use a mask
    2. Replace it with a sentinel value (e.g., -999)
    3. Do nothing (decide later)
    
[back to top](#top)
```{r, eval=FALSE}
elec_data[['2008']]
```

### What are missing values in R? 
- There are three (four) types.
    - `NULL`: Non existence.
    - `NaN` (Not a Number): A missing numerical value (double).
    - `NA` (Not Available): A missing value (logical).
    - (`Inf`: An infinite number.)
- Print vectors in the below example.
- What happens if you aggregate all items in each vector?
    - Try `sum()` and `sum( , na.rm=TRUE)`. This `na.rm=TRUE` option ignores `NaN` and `NA`.
```{r, eval=FALSE}
(vec1 <- c(1, 2, 3, NA))
(vec2 <- c(1, 2, 3, NaN))
(vec3 <- c(1, 2, 3, NULL))
(vec4 <- c(1, 2, 3, Inf))
```

### Handling missing values
- Useful functions:
    - Checking: `is.na()` and `!is.na()`.
    - Also, you can combine them with `any()` or `all()`.
- Print `vec1` and `df1`.
- Replace missing values with zeros in `vec1` or `df1`.
    - Use `myvec[is.na(myvec)] <- 0` and `mydf[is.na(mydf)] <- 0`.
- Drop missing values in `vec1` or `df1`.
    - Use `myvec <- myvec[!is.na(myvec)]` and `mydf <- mydf[!is.na(mydf)]`.
```{r, eval=FALSE}
vec1 <- c(1.0, NaN, 3.0, NA, NULL)
df1 <- as.data.frame(matrix(c(1.0, NA, 3.0, NA, 5.0, NaN, 7.0, 8.0, NaN, NaN, NaN, NaN, NA, NA, NA, NA), ncol=4))
```

- To drop columns whose rows are all missing, you can use `colSums`.
- To drop rows whose rows are all missing, you can use `rowSums`.
- Print `df2` and `df3`.
```{r, eval=FALSE}
df1 <- as.data.frame(matrix(c(1.0, NA, 3.0, NA, 5.0, NaN, 7.0, 8.0, NaN, NaN, NaN, NaN, NA, NA, NA, NA), ncol=4))
print(df1)
print(colSums(is.na(df1)))
df2 <- df1[, colSums(is.na(df1)) != nrow(df1)]
print(rowSums(is.na(df1)))
df3 <- df1[rowSums(is.na(df1)) != ncol(df1), ]
```

## 3. Selecting Columns<a id='sec3'></a>
- Data may contain some redundant columns that will never be used in analysis. We will drop such columns to reduce the data size.
- Candidates:
    1. A column whose values are all missing
    2. A column whose information is not important
- You should be very careful about selecting columns.
    - You may not use those columns in the current project but may use them in another project.
- Let's check `elec_data[['2008']]`.
    
[back to top](#top)
```{r, eval=FALSE}
elec_data[['2008']]
```

- All columns look good. What about `temp_data[['2008']]`?
```{r, eval=FALSE}
head(temp_data[['2008']])
# unique(temp_data[['2008']]['Parameter.Name']) # unique() prints unique entries
```

- Let's drop `Parameter.Name`, `Sample.Duration`, `Pollutant.Standard`, `Units.of.Measure`, and `AQI`.
- First, remove columns that are all missing.
```{r, eval=FALSE}
for (year in seq(2008,2014,by=2)) {
    temp_data[[paste0(year, '_keep')]] <- temp_data[[as.character(year)]][, colSums(is.na(temp_data[[as.character(year)]])) != nrow(temp_data[[as.character(year)]])]    
}
```

- Next, drop the remaining columns.
```{r, eval=FALSE}
for (year in seq(2008,2014,by=2)) {
    temp_data[[paste0(year, '_keep')]] <- temp_data[[paste0(year, '_keep')]][, !names(temp_data[[paste0(year, '_keep')]]) %in% c('Parameter.Name', 'Sample.Duration', 'Units.of.Measure')]
}
```

### Changing Column Names
- If column names use uppercases and space(s), are very long, or are written in any other non-generic format, we need to change them.
- Check the column names in `temp_data[['2008_keep']]`.
```{r, eval=FALSE}
colnames(temp_data[['2008_keep']])
```

- I suggest that we (a) replace periods with underscores and (b) change uppercases to lowercases.
    - To replace spaces with underscores, you can use `gsub()`.
    - To change to lowercases, you can use `tolower()`. If you want to use uppercases, use `toupper()`.
- Run the following code and check the column names in `temp_data[['2008_keep']]`.
```{r, eval=FALSE}
for (year in seq(2008,2014,by=2)) {
    colnames(temp_data[[paste0(year, '_keep')]]) <- gsub("\\.", "_", colnames(temp_data[[paste0(year, '_keep')]]))
    colnames(temp_data[[paste0(year, '_keep')]]) <- tolower(colnames(temp_data[[paste0(year, '_keep')]]))
}
colnames(temp_data[['2008_keep']])
```

## 4. Selecting Rows<a id='sec4'></a>
- Next, let's move on to rows.
- We need to decide which rows to keep in the final data.
- Candidates:
    - Rows that have missing values for important variables such as id
    - Multiple rows per unit of observation
    
[back to top](#top)

- First, let's look at `elec_data[['2008']]` again. Recall that all election results are missing for some rows.
```{r, eval=FALSE}
elec_data[['2008']]
```

- Let's drop such rows.
- Run the following code.
```{r, eval=FALSE}
for (year in seq(2008,2014,by=2)) {
    elec_data[[paste0(year, '_keep')]] <- elec_data[[as.character(year)]][rowSums(is.na(elec_data[[as.character(year)]][c('gelec_dem', 'gelec_rep', 'gelec_oth')])) != ncol(elec_data[[as.character(year)]][c('gelec_dem', 'gelec_rep', 'gelec_oth')]), ]
}
elec_data[['2008_keep']]
```

- Next, print `'date_local'` column for `"New York"` using `'state_name'` in `temp_data[['2008_keep']]`.
    - What does this column mean?
    - Should we drop all the dates that are not needed for the analysis? Election Days are:
        - November 4th, 2008, November 2nd, 2010, November 6th, 2012, November 8th, 2014
```{r, eval=FALSE}
temp_data[['2008_keep']][temp_data[['2008_keep']]['state_name'] == "New York", 'date_local']
```

- I suggest we keep all dates. 
    - The reason: We may use daily temperature on non-Election days as well.
- Next, print `temp_data[['2008_keep']]` for `"New York"` and `"2008-11-04"`, i.e., Election Day. We know that there are multiple entries.
```{r, eval=FALSE}
temp_data[['2008_keep']][(temp_data[['2008_keep']]['state_name'] == "New York") & (temp_data[['2008_keep']]['date_local'] == "2008-11-04"), ]
```

- What should we do? There are several options. What are the pros and cons of each method?
    1. Aggregate data
    2. Reshape data
    3. Delete some observations
    4. Do nothing (decide later)

- There is no right answer!
- Let's take the fourth option for now. Two reasons: 
    - There is no point of losing rich information in data (regarding 1 and 3). 
    - There will be many missing values if we reshape them (regarding 2).
    
## 5. Treating Other Values<a id='sec5'></a>
- It could often be the case that you need to modify entries in a dataset "by hand".
    - Example: Some data entries are incorrect.
- The bottom line: Save code for whatever you did for original data, if possible.
    
[back to top](#top)

## 6. Saving Data<a id='sec6'></a>
- You can save data in any format. However, the most preferred one is csv.
- To save data as a csv file, use `write.table()`.
    - `sep=","` option should be added for csv files.
    - `na=""` means that missing values will be produced as blank cells.
    - `row.names=FALSE` means that row names will not be written as a character vector of row names.
- Use `write.xlsx()` in `readxl` package to save in excel format.
    
[back to top](#top)
```{r, eval=FALSE}
for (year in seq(2008,2014,by=2)) {
    write.table(elec_data[[paste0(year, '_keep')]], file=paste0('us_senate_R_', year, '.csv'), sep=",", na="", row.names=FALSE)
    write.table(temp_data[[paste0(year, '_keep')]], file=paste0('us_daily_temp_R_', year, '.csv'), sep=",", na="", row.names=FALSE)   
}
```








